### 2013-11-10 src2/hl/paperThesis.tex
update

\begin{small} 
\begin{verbatim} 
reverses :: DPull a -> DPush Grid a
reverses arr = pConcatMap (pJoin . reverseLocal)
                          (splitUp 256 arr)

reverses' :: MemoryOps a => DPull a -> DPush Grid a
reverses' arr = pConcatMap (pJoin . reverseLocal')
                           (splitUp 256 arr) 
\end{verbatim} 
\end{small}

\noindent
The {\tt pConcatMap} combinator has the following type: 
%takes a local computation that results in a push array.
\begin{small}
\begin{Verbatim}[samepage=true] 
pConcatMap :: ASize l 
            => (a1 -> SPush t a) 
            -> Pull l a1 
            -> Push (Step t) l a
\end{Verbatim}
\end{small}

{\tt pConcatMap} takes a computation resulting in a push array at a specific 
level of the GPU hierarchy, {\tt t}, and a pull array. The computation 
is applied to each element of the pull array and the results are concatenated 
into a push array at the level above {\tt t}. Here, this is used to apply a block 
level computation, giving a grid computation. The {\tt pJoin} function 
has type \verb!Program t (Push t s a) -> Push t s a!; it creates a new push array 
that computes the same results as the input computation.

@@

\begin{small} 
\begin{verbatim} 
largeReverse :: DPull b
              -> Push Grid EWord32 b
largeReverse arr =
  pConcat $ reverse $ fmap (pJoin . reverseLocal)
                      (splitUp 256 arr) 
\end{verbatim} 
\end{small}

@@

\begin{small} 
\begin{verbatim} 
iters :: EWord32 
         -> EWord32 
         -> TProgram (SPush Thread EWord8)
iters bid tid =
  return $ fmap extract 
         $ seqUntil (f bid tid) cond  (0,0,1)
  where
    extract (_,_,c) = ((w32ToW8 c) `mod` 16) * 16
\end{verbatim}
\end{small}

@@

\begin{small} 
\begin{verbatim} 
genRect :: MemoryOps b
           => EWord32
           -> Word32
           -> (EWord32 -> EWord32 -> SPush Thread b)
           -> DPush Grid b 
genRect bs ts p = generate bs $
                  \bid -> generate ts $ p bid 
\end{verbatim}
\end{small} %$

The largest possible image that can be obtained using this particular 
method is 1024$\times$1024. This is because of the maximum threads per block 
limitation of current GPUs. The code below generates a 512$\times$512 element 
image. 
\begin{small} 
\begin{verbatim}
mandel = genRect 512 512 (\b -> pJoin . iters b)
\end{verbatim}
\end{small}

@@

\begin{small} 
\begin{Verbatim}[samepage = true] 
mapRed1 :: MemoryOps a 
              => (a -> a -> a) 
              -> DPull (SPull a) 
              -> DPush Grid a
mapRed1 f = pConcatMap $ pJoin . red1 f
\end{Verbatim}
\end{small} %$

@@

\begin{small}
\begin{Verbatim}[samepage = true] 
red4 :: MemoryOps a
           => (a -> a -> a)
           -> SPull a
           -> BProgram (SPush Block a)
red4 f arr =
  do
    arr' <- force $  
            pConcatMap (return . seqReduce f) 
                       (splitUp 8 arr)
    red3 f arr' 
\end{Verbatim}
\end{small}%$

@@

\begin{small} 
\begin{Verbatim}[samepage = true] 
red5 :: MemoryOps a
           => (a -> a -> a)
           -> SPull a
           -> BProgram (SPush Block a)
red5 f arr =
  do
    arr' <- force $  
            pConcatMap (return . seqReduce f)
                       (coalesce 8 arr)
    red3 f arr'

>>>>

\begin{small} 
\begin{verbatim} 
reverses :: DPull a -> DPush Grid a
reverses arr = pConcatMap reverseLocal
                          (splitUp 256 arr)

reverses' :: MemoryOps a => DPull a -> DPush Grid a
reverses' arr = pConcatMap reverseLocal'
                           (splitUp 256 arr) 
\end{verbatim} 
\end{small}

\noindent
The {\tt pConcatMap} combinator is overloaded; here it is applied 
at the following type: 
%takes a local computation that results in a push array.
\begin{small}
\begin{Verbatim}[samepage=true] 
pConcatMap :: ASize l 
            => (a1 -> Program t (SPush t a))
            -> Pull l a1 
            -> Push (Step t) l a
\end{Verbatim}
\end{small}

{\tt pConcatMap} takes a computation resulting in a push array at a specific 
level of the GPU hierarchy, {\tt t}, and a pull array. The computation 
is applied to each element of the pull array and the results are concatenated 
into a push array at the level above {\tt t}. Here, this is used to apply a block 
level computation, giving a grid computation.

@@

\begin{small} 
\begin{verbatim} 
largeReverse :: DPull b
              -> Push Grid EWord32 b
largeReverse arr =
  pConcat $ reverse $ fmap reverseLocal
                      (splitUp 256 arr) 
\end{verbatim} 
\end{small}

@@

\begin{small} 
\begin{verbatim} 
iters :: EWord32 -> EWord32 -> SPush Thread EWord8
iters bid tid = fmap extract $
                seqUntil (f bid tid) cond (0,0,1)
  where
    extract (_,_,c) = ((w32ToW8 c) `mod` 16) * 16
\end{verbatim}
\end{small}

@@

\begin{small} 
\begin{verbatim} 
genRect :: EWord32
          -> Word32
          -> (EWord32 -> EWord32 -> SPush Thread b)
          -> DPush Grid b 
genRect bs ts p = generate bs $
                  \bid -> generate ts $ p bid 
\end{verbatim}
\end{small} %$

The largest possible image that can be obtained using this particular 
method is 1024$\times$1024. This is because of the maximum threads per block 
limitation of current GPUs. The code below generates a 512$\times$512 element 
image. 
\begin{small} 
\begin{verbatim}
mandel = genRect 512 512 iters
\end{verbatim}
\end{small}

@@

\begin{small} 
\begin{Verbatim}[samepage = true] 
mapRed1 :: MemoryOps a 
              => (a -> a -> a) 
              -> DPull (SPull a) 
              -> DPush Grid a
mapRed1 f = pConcatMap $ red1 f
\end{Verbatim}
\end{small} %$

@@

\begin{small}
\begin{Verbatim}[samepage = true] 
red4 :: MemoryOps a
           => (a -> a -> a)
           -> SPull a
           -> BProgram (SPush Block a)
red4 f arr =
  do
    arr' <- force $  
            pConcatMap (seqReduce f) 
                       (splitUp 8 arr)
    red3 f arr' 
\end{Verbatim}
\end{small}%$

@@

\begin{small} 
\begin{Verbatim}[samepage = true] 
red5 :: MemoryOps a
           => (a -> a -> a)
           -> SPull a
           -> BProgram (SPush Block a)
red5 f arr =
  do
    arr' <- force $  
            pConcatMap (seqReduce f)
                       (coalesce 8 arr)
    red3 f arr'

### 2013-11-10 src2/thesis.tex
update

Computers today are becoming more and more parallel. General purpose processors (CPUs) 
have multiple processing cores and Single Instruction Multiple Data (SIMD) units for 
data-parallelism. Graphics processors (GPUs) brings massive parallelism at the cost of 
being harder to program than CPUs. This thesis applies embedded language methodology 
to data-parallel programming. Two embedded languages are presented, Obsidian for general 
purpose GPU programming and EmbArBB for data-parallel programming across platforms.

Each generation of CPU and GPU bringing more parallel resources. The question of how to write 
programs for these processors efficiently arise. We are after efficiency both in programmer 
productivity and in application performance. Using embedded languages allows us to experiment
with what abstractions to present to the programmer at relatively little effort.

@@

In cases where the programmer is not interested in low-level details 
or wants the program to run over a range of hardware a higher level language can be used.
EmbArBB is a Haskell embedding or the Intel ArBB system. EmbArBB relies on the ArBB system 
to generate code (via a Just-In-Time compiler) to a range of hardware.

EmbArBB embeds a preexisting library for data-parallelism into Haskell and we obtain 
very good performance at little implementation effort. This performance comes from the 
expertise and effort put into the ArBB system and that we get for free. Embedding ArBB is 
a way to provide these benefits to the Haskell programmer and a way to increase usefulness
of an existing system by opening it up to a wider audience. Obsidian is very different, it 
is not based on a set of high-level parallel primitives. The Obsidian programmer 
can implement these primitives in different ways, select the best one and then  abstract 
upon that. We have obtained very good performance in case studies involving reductions.

@@

I would like to thank my supervisor Mary Sheeran. She is a very suportive, motivating and 
understanding supervisor. I also thank my co-supervisor Koen Claessen for all the great advice 
and tips he has provided over he years. I cannot imagine that this would have worked with 
any other supervisors. Thanks to Josef Svenningsson, a late addition to my troop of 
supervisors. Working with Josef has been a very fun and rewarding experience.

@@

\begin{small} 
\begin{Verbatim}[samepage=true]
vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
vecAddG i1 i2  =
  pConcat $
  zipWith vecAdd
          (splitUp 32 i1)
          (splitUp 32 i2)  
\end{Verbatim}
\end{small} % $

@@

\begin{small}
\begin{Verbatim}[samepage=true]
vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
vecAddG i1 i2  =
  pConcat $
  zipWith (zipWith (+)) 
          (splitUp 32 i1)
          (splitUp 32 i2)
\end{Verbatim} 
\end{small} % $

>>>>

Computers today are becoming more and more parallel. General purpose processors (CPUs) 
have multiple processing cores and Single Instruction Multiple Data (SIMD) units for 
data-parallelism. Graphics processors (GPUs) bring massive parallelism at the cost of 
being harder to program than CPUs. This thesis applies embedded language methodology 
to data-parallel programming. Two embedded languages are presented, Obsidian for general 
purpose GPU programming and EmbArBB for data-parallel programming across platforms.

CPUs and GPUs get more parallel resources with each new generation. The question of how to 
program these processors efficiently arises. We are after efficiency both in programmer 
productivity and in application performance. Using embedded languages allows us to experiment
with what abstractions to present to the programmer at relatively little effort.

@@

In cases where the programmer is not interested in low-level details 
or wants the program to run over a range of hardware, a higher level language can be used.
EmbArBB is a Haskell embedding or the Intel ArBB system. EmbArBB relies on the ArBB system 
to generate code (via a Just-In-Time compiler) to a range of hardware.

EmbArBB embeds a preexisting library for data-parallelism into Haskell and we obtain 
very good performance at little implementation effort. This performance comes from the 
expertise and effort put into the ArBB system and that we get for free. Embedding ArBB is 
a way to provide these benefits to the Haskell programmer and a way to increase usefulness
of an existing system by opening it up to a wider audience. Obsidian is very different; it 
is not based on a set of high-level parallel primitives. The Obsidian programmer 
can implement these primitives in different ways and then select the best one. 
We have obtained very good performance in case studies involving reductions.

@@

I would like to thank my supervisor Mary Sheeran. She is a very suportive, motivating and 
understanding supervisor. I also thank my co-supervisor Koen Claessen for all the great advice 
and tips he has provided over the years. I cannot imagine that this would have worked with 
any other supervisors. Thanks to Josef Svenningsson, a late addition to my troop of 
supervisors. Working with Josef has been fun and rewarding.

@@

\begin{small} 
\begin{Verbatim}[samepage=true]
vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
vecAddG i1 i2  =
  pConcat $
  fmap push $ 
  zipWith vecAdd
          (splitUp 32 i1)
          (splitUp 32 i2)  
\end{Verbatim}
\end{small}

@@

\begin{small}
\begin{Verbatim}[samepage=true]
vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
vecAddG i1 i2  =
  pConcat $
  fmap push $
  zipWith (zipWith (+)) 
          (splitUp 32 i1)
          (splitUp 32 i2)
\end{Verbatim} 
\end{small}
